% Building Multiplayer Unreal Applications with Autonomous AI Assistance
% Michael Clark — Are-Self Research
\documentclass[man,12pt,floatsintext]{apa7}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{listings}
\usepackage{xcolor}

\definecolor{codebg}{rgb}{0.95,0.95,0.97}
\definecolor{codegreen}{rgb}{0.2,0.6,0.2}
\definecolor{codegray}{rgb}{0.5,0.5,0.5}
\definecolor{codepurple}{rgb}{0.58,0,0.82}

\lstset{
  backgroundcolor=\color{codebg},
  basicstyle=\ttfamily\small,
  breaklines=true,
  captionpos=b,
  commentstyle=\color{codegreen},
  keywordstyle=\color{codepurple},
  numberstyle=\tiny\color{codegray},
  stringstyle=\color{codegreen},
  frame=single,
  numbers=left,
}

\usepackage[style=apa,backend=biber]{biblatex}
\addbibresource{../../templates/references.bib}

\title{Building Multiplayer Unreal Applications with Autonomous AI Assistance}
\shorttitle{MULTIPLAYER UNREAL + AUTONOMOUS AI}
\author{Michael Clark}
\affiliation{Independent Researcher \\ scipraxian}

\abstract{
Game development with Unreal Engine is one of the most complex software engineering
disciplines, requiring expertise in C++, Blueprints, networking, rendering, physics,
and asset pipelines. For students and small teams without enterprise resources, this
complexity is a barrier to entry. We describe how Are-Self's autonomous AI agents assist
in building a multiplayer Unreal Engine application by managing build pipelines,
analyzing shader compilation, automating save game management, coordinating multi-platform
testing, and providing reasoning-assisted debugging. The AI operates as a project manager
and build engineer, using Are-Self's neural pathway system for workflow automation and
the Frontal Lobe for intelligent decision-making about build failures, performance
regressions, and deployment sequencing. We present lessons learned from developing a
multiplayer Unreal project with Are-Self as an integral part of the development workflow.
}

\keywords{Unreal Engine, game development, autonomous AI, multiplayer, build automation,
AI-assisted development}

\begin{document}
\maketitle

\section{Introduction}

Unreal Engine is among the most powerful game engines available, offered free for
educational use and indie development. However, its power comes with extraordinary
complexity: a multiplayer Unreal project involves C++ and Blueprint programming,
dedicated server configuration, replication networking, shader compilation (Pipeline
State Object caching), asset cooking, platform-specific packaging, and multi-stage
testing.

Professional studios employ dedicated build engineers and DevOps teams to manage this
complexity. Students, educators, and small indie teams do not have this luxury.

Are-Self was developed alongside an Unreal Engine multiplayer project, and many of its
neural pathways were built specifically to automate game development workflows. This paper
describes the symbiotic relationship between the AI orchestration system and the game
project it helps build.

\section{The Unreal Development Workflow}

\subsection{Build Pipeline Complexity}

% TODO: Describe the multi-stage Unreal build pipeline
% - C++ compilation
% - Blueprint compilation
% - Shader compilation and PSO caching
% - Asset cooking
% - Packaging (client + dedicated server)
% - Multi-platform considerations

\subsection{Multiplayer-Specific Challenges}

% TODO: Describe multiplayer-specific issues
% - Dedicated server vs. listen server
% - Replication debugging
% - Network prediction and reconciliation
% - Session management
% - Cross-platform multiplayer testing

\section{Are-Self Integration Points}

\subsection{Neural Pathways for Build Automation}

We describe each Are-Self pathway used in the Unreal development workflow:

\begin{enumerate}
  \item \textbf{Pre-Release}: Version metadata $\rightarrow$ save backup $\rightarrow$
        build $\rightarrow$ test run (5-node pathway)
  \item \textbf{Compile Shaders}: PSO cache generation across target platforms
  \item \textbf{Deploy}: Multi-environment deployment with rollback on failure
  \item \textbf{Staging Run}: Quick build-and-test for development iteration
  \item \textbf{Dump Scripts / Record PSOs}: Data extraction utilities
\end{enumerate}

\subsection{AI-Assisted Debugging}

% TODO: How the Frontal Lobe assists with build failures
% - Build log analysis
% - Shader compilation error diagnosis
% - Replication bug reasoning
% - Suggested fixes via mcp_respond_to_user

\subsection{Project Management}

% TODO: How the Prefrontal Cortex manages game development tasks
% - Epics for major features
% - Stories for implementation chunks
% - Tasks for individual work items
% - AI-driven task decomposition and estimation

\subsection{Memory-Assisted Development}

% TODO: How the Hippocampus stores development knowledge
% - Build configuration engrams
% - Known issues and workarounds
% - Platform-specific gotchas
% - Carried across sessions

\section{Case Study: A Multiplayer Unreal Project}

% TODO: Detailed case study
% - Project description (multiplayer game type)
% - Development timeline with Are-Self integration
% - Specific examples of AI assistance
% - Quantitative metrics (build success rate, time savings)

\section{Lessons Learned}

% TODO: Key insights
% - Where AI assistance was most valuable
% - Where it fell short
% - How pathways evolved over development
% - The value of persistent memory across sessions

\section{Discussion}

\subsection{Accessibility Impact}

Are-Self lowers the barrier for Unreal Engine multiplayer development by automating the
most complex operational aspects. A student can focus on game design and programming while
the AI handles build orchestration, shader management, and deployment.

\subsection{Limitations}

The AI cannot directly modify C++ or Blueprint code (tool limitation). Complex
networking bugs may exceed the reasoning capabilities of 7B local models. Real-time
game testing still requires human players.

\section{Conclusion}

We have demonstrated that an autonomous AI orchestration system can serve as an effective
build engineer and project manager for multiplayer Unreal Engine development. By combining
neural pathway automation with AI-assisted reasoning, Are-Self makes complex game
development workflows accessible to students and small teams who lack dedicated DevOps
resources.

\printbibliography

\end{document}
