% CI/CD Sovereignty: Self-Managed Build and Deployment Pipelines for Local AI Systems
% Michael Clark — Are-Self Research
\documentclass[man,12pt,floatsintext]{apa7}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{listings}
\usepackage{xcolor}

\definecolor{codebg}{rgb}{0.95,0.95,0.97}
\definecolor{codegreen}{rgb}{0.2,0.6,0.2}
\definecolor{codegray}{rgb}{0.5,0.5,0.5}
\definecolor{codepurple}{rgb}{0.58,0,0.82}

\lstset{
  backgroundcolor=\color{codebg},
  basicstyle=\ttfamily\small,
  breaklines=true,
  captionpos=b,
  commentstyle=\color{codegreen},
  keywordstyle=\color{codepurple},
  numberstyle=\tiny\color{codegray},
  stringstyle=\color{codegreen},
  frame=single,
  numbers=left,
}

\usepackage[style=apa,backend=biber]{biblatex}
\addbibresource{../../templates/references.bib}

\title{CI/CD Sovereignty: Self-Managed Build and Deployment Pipelines for Local AI Systems}
\shorttitle{CI/CD SOVEREIGNTY}
\author{Michael Clark}
\affiliation{Independent Researcher \\ scipraxian}

\abstract{
Cloud-hosted CI/CD services (GitHub Actions, GitLab CI, CircleCI) are effective for
open-source projects but introduce external dependencies that conflict with Are-Self's
local-first, privacy-preserving design philosophy. We describe how Are-Self's own
neural pathway system can serve as a CI/CD orchestrator, executing build, test, and
deployment pipelines as spike trains through the Central Nervous System. This approach
eliminates cloud CI dependencies, enables AI-assisted pipeline management (the Frontal
Lobe can reason about build failures), and maintains the system's sovereignty over its
own infrastructure. We demonstrate pipelines for Unreal Engine game builds, shader
compilation, automated testing, and multi-environment deployment---all orchestrated by
the same spike-train architecture that drives Are-Self's reasoning sessions.
}

\keywords{CI/CD, local infrastructure, build automation, neural pathways, sovereignty,
self-hosted, Unreal Engine}

\begin{document}
\maketitle

\section{Introduction}

Are-Self's target users---students, educators, nonprofits---often lack access to paid
CI/CD services or the expertise to configure complex cloud pipelines. Yet their projects
(game development, educational tools, research code) benefit enormously from automated
build and deployment.

Are-Self already contains a generic workflow execution engine: the Central Nervous System,
which fires spike trains through directed graphs of effector nodes. We observe that CI/CD
pipelines are directed graphs of tasks with conditional routing---exactly the computation
the CNS already performs.

This paper describes how Are-Self's existing infrastructure doubles as a CI/CD system,
with the unique advantage that an AI reasoning agent (the Frontal Lobe) can inspect build
failures, suggest fixes, and adapt pipelines in real time.

\section{Related Work}

\subsection{Cloud CI/CD}

% TODO: GitHub Actions, GitLab CI, Jenkins
% - Strengths: ecosystem, community actions, scalability
% - Weaknesses: external dependency, cost, privacy implications

\subsection{Self-Hosted CI/CD}

% TODO: Jenkins, Drone, Woodpecker, Gitea Actions
% - Strengths: control, privacy
% - Weaknesses: maintenance burden, no AI assistance

\subsection{AI-Assisted DevOps}

% TODO: GitHub Copilot for CI, AI-powered pipeline optimization
% - Cloud-dependent, not local

\section{Architecture}

\subsection{Pipelines as Neural Pathways}

A CI/CD pipeline maps directly to a NeuralPathway:

\begin{itemize}
  \item Pipeline stages $\rightarrow$ Neurons with Effectors
  \item Stage ordering $\rightarrow$ Axon connections (FLOW type)
  \item Conditional steps $\rightarrow$ SUCCESS/FAILURE axon routing
  \item Error handling $\rightarrow$ Gate nodes (condition checking) + Retry nodes
  \item Environment variables $\rightarrow$ Context Variables resolved per Environment
\end{itemize}

\subsection{Example: Unreal Engine Pre-Release Pipeline}

% TODO: Walk through the Pre-Release pathway as a CI/CD pipeline
% - Begin Play → Version Metadata → Backup SaveGames → Pre-Release Build → Pre-Release Run
% - Each node as an effector wrapping a shell command
% - Environment variable resolution (project root, build config)
% - FAILURE routing for error handling

\subsection{AI-Assisted Pipeline Management}

When a build fails, the spike's FAILURE axon can route to a Frontal Lobe effector node.
The reasoning agent receives the build log in its blackboard context, reasons about the
failure, and can:

\begin{enumerate}
  \item Diagnose the root cause
  \item Suggest a fix to the user via \texttt{mcp\_respond\_to\_user()}
  \item Modify pipeline parameters for a retry
  \item File a PFC task for manual follow-up
\end{enumerate}

\subsection{Multi-Environment Deployment}

Are-Self's Environment system naturally supports multi-environment deployment:

\begin{itemize}
  \item Development, Staging, Production as separate Environments
  \item Same pathway, different context variables per environment
  \item Environment-scoped effector configurations
\end{itemize}

\section{Implementation}

% TODO: Technical details
% - Effector wrapping shell commands (Executable model)
% - Blackboard propagation through pipeline stages
% - Telemetry collection (build times, success rates via sparklines)
% - Distribution modes for parallel builds across PNS agents

\section{Evaluation}

% TODO: Real-world pipeline case studies
% - Unreal Engine game build (compile, cook, package, deploy)
% - Django test suite execution
% - Multi-platform shader compilation (PSO cache)
% - Documentation site deployment

\section{Discussion}

\subsection{Sovereignty Benefits}

No external service has access to source code, build artifacts, or deployment credentials.
The entire pipeline runs on the same hardware as the AI reasoning engine.

\subsection{Limitations}

Complex CI/CD features (matrix builds, container orchestration, artifact caching) require
additional development beyond basic pathway execution. The current system lacks native
Git integration for trigger-on-push workflows.

\section{Conclusion}

Are-Self's neural pathway system provides a natural substrate for CI/CD pipelines,
eliminating cloud dependencies while adding AI-assisted failure diagnosis. For the
target audience of students and small organizations, this means automated builds and
deployments without paid services, cloud accounts, or CI/CD expertise.

\printbibliography

\end{document}
